---
title: "Figure 7c-f: re-cluster CD4 T cells from Wu et al, find CD4 Trms and correlate with cDC1s"
author: "Ignacio Heras"
documentclass: article
output:
  html_document: 
    number_sections: yes
    self_contained: no
    theme: united
    toc: yes
    toc_depth: 5  
    toc_float: true
    collapsed: no
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
urlcolor: blue
classoption: a4paper
---

```{r setup, include=FALSE}
## packages
suppressMessages(library("Matrix"))
suppressMessages(library("ggplot2"))
suppressMessages(library("ggrepel"))
suppressMessages(library("ggpubr"))
suppressMessages(library("ggridges"))
suppressMessages(library("dplyr"))
suppressMessages(library("tidyr"))
suppressMessages(library("readr"))
suppressMessages(library("biomaRt"))
suppressMessages(library("Seurat"))
suppressMessages(library("scater"))
suppressMessages(library("org.Mm.eg.db"))
suppressMessages(library("ggvenn"))
suppressMessages(library("gtools"))
suppressMessages(library("viridis"))
suppressMessages(library("plotly"))
suppressMessages(library("Matrix.utils"))
library(colorRamp2)
suppressMessages(library("SingleCellExperiment"))

## paths
projectPath <- here::here()
dataPath <- here::here("data")
metadataPath <- here::here("metadata")
sourceCode <- here::here("src")
objectsPath <- here::here("trial5")
processedPath <- here::here("trial5/processed_data")
reportsPath <- here::here("reports")
plotsPath <- here::here(reportsPath, "plots")
prefix <- "trial5"

## scr code
source(file.path(sourceCode, "helperFunctions.R"))

## rmarkdown settings
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.path = file.path(reportsPath, "plots/"),
  fig.pos = "H",
  message = FALSE,
  warning = FALSE,
  comment = NA,
  dev = "png",
  dpi = 300
)

## global options
options(encoding = 'UTF-8')
```


# Loading data

```{r}
## metadata
cellsMetadata <- read.csv(
  file.path(dataPath, "SingleCellPortal", "metadata", "Whole_miniatlas_meta.csv")
)[-1, ]
rownames(cellsMetadata) <- cellsMetadata$NAME
```


```{r}
if (file.exists(file.path(objectsPath, paste(prefix, "seurat.BC.rds", sep = ".")))) {
  seurat.BC <- readRDS(
    file.path(objectsPath, paste(prefix, "seurat.BC.rds", sep = "."))
  )
} else {
  ## reading raw counts
  scDataDir <- file.path(dataPath, "SingleCellPortal", "BrCa_Atlas_Count_out")
  rawData <- Read10X(data.dir = scDataDir)
  if (!all(cellsMetadata$NAME == colnames(rawData))) {
    stop("Error: problem with cells metadata")
  }
  seurat.BC <- CreateSeuratObject(
    counts = rawData, 
    meta.data = cellsMetadata,
    names.delim = "-",
    project = "BreastCancer_NatGen_2021"
  )  %>% NormalizeData(
    normalization.method = "LogNormalize", verbose = TRUE
  )
  saveRDS(
    seurat.BC, file.path(objectsPath, paste(prefix, "seurat.BC.rds", sep = "."))
  )
}

```
# Normalization and scaling

```{r}
seurat.BC <- FindVariableFeatures(
  seurat.BC, selection.method = "vst"
)
## visualization
top10 <- head(VariableFeatures(seurat.BC))
varGenes <- VariableFeatures(seurat.BC)
# plot variable features with and without labels
plot <- VariableFeaturePlot(seurat.BC)
LabelPoints(
  plot = plot, points = top10, 
  repel = TRUE, xnudge = 0, ynudge = 0
)
```


```{r}
if (file.exists(file.path(objectsPath, paste(prefix, "seurat.BC.scal.rds", sep = ".")))) {
  seurat.BC <- readRDS(
    file = file.path(
      objectsPath, paste(prefix, "seurat.BC.scal.rds", sep = ".")
    )
  )
} else {
  seurat.BC <- ScaleData(
    object = seurat.BC, 
    do.scale = TRUE,
    do.center = TRUE,
    features = varGenes,
    display.progress = TRUE
  ) %>% RunPCA(
    features = varGenes,
    ndims.print = 1:5,
    nfeatures.print = 30,
    npcs = 100
  )
  saveRDS(
    seurat.BC, 
    file = file.path(objectsPath, paste(prefix, "seurat.BC.scal.rds", sep = "."))
  )
}
```

```{r, eval = FALSE}
list.seurat.BCs <- SplitObject(seurat.BC, split.by = "ident")
```

Taken from (<https://github.com/Swarbricklab-code/BrCa_cell_atlas/blob/main/seurat_data_integration/config/scripts/seurat_CCA_HPC_processing.R>).

```{r, eval = FALSE}
num_cc_to_compute <- 30
num_genes_for_int_CCA <- 2000
integrate_all_genes_by_intersect <- FALSE
number_of_PCs_to_compute <- 100
run_parallel <- TRUE
number_of_objects <- length(list.seurat.BCs)
```


```{r, eval = FALSE}
# Find integration anchors
if (run_parallel) {
  plan("multiprocess", workers = temp_run_parallel_cores)
  options(future.globals.maxSize = 10 * 1024^3)
}

temp_anchor_start <- Sys.time()
seuratAnchors <- FindIntegrationAnchors(
  object.list = list.seurat.BCs, 
  dims = 1:num_cc_to_compute,
  anchor.features = num_genes_for_int_CCA
)
temp_anchor_finish <- Sys.time()

# intersection
if (integrate_all_genes_by_intersect) {
  for (i in c(1:length(temp_sample_list))) {
    temp_genes <- rownames(temp_sample_list[[i]])
    n <- paste0("temp_genes_sample_", i)
    assign(n, temp_genes)
  }
  temp_genes_list <- mget(ls(pattern = "temp_genes_sample_*"))
  temp_genes_interesect <- Reduce(intersect, temp_genes_list)
} else {
  temp_integration_start <- Sys.time()
  seurat.BC.int <- IntegrateData(
    anchorset = seuratAnchors, 
    dims = 1:temp_num_cc_compute
  )
  temp_integration_finish <- Sys.time()
}
 
DefaultAssay(object = seurat.BC.int) <- "integrated"

# scale data
seurat.BC.int <- ScaleData(
  object = seurat.BC.int, verbose = TRUE
)

# PCA
seurat.BC.int <- RunPCA(
  object = seurat.BC.int, npcs = number_of_PCs_to_compute, 
  verbose = TRUE
)

```

# Load object

```{r}
#Inspired from last comment on https://github.com/satijalab/seurat/issues/1883
seurat.BC <- readRDS(file.path(objectsPath, paste(prefix, "seurat.BC.int.rds", sep = ".")))
  
#Compile clusters for Diego: 
seurat.BC@meta.data <- mutate(
  seurat.BC@meta.data,
  celltype_custom = case_when(
    celltype_subset %in% c("T_cells_c2_CD4+_T-regs_FOXP3") ~ "CD4 Tregs",
    celltype_subset %in% c("T_cells_c3_CD4+_Tfh_CXCL13", "T_cells_c0_CD4+_CCR7", "T_cells_c1_CD4+_IL7R") ~ "CD4 TCs",
    celltype_subset %in% c("T_cells_c4_CD8+_ZFP36", "T_cells_c5_CD8+_GZMK", "T_cells_c7_CD8+_IFNG", "T_cells_c8_CD8+_LAG3") ~ "CD8 TCs",
    celltype_subset %in% c("T_cells_c9_NK_cells_AREG", "T_cells_c10_NKT_cells_FCGR3A") ~ "NK",
    celltype_subset %in% c("Myeloid_c1_LAM1_FABP5", "Myeloid_c10_Macrophage_1_EGR1", "Myeloid_c2_LAM2_APOE", "Myeloid_c9_Macrophage_2_CXCL10") ~ "MFs",
    celltype_subset %in% c("Myeloid_c7_Monocyte_3_FCGR3A", "Myeloid_c8_Monocyte_2_S100A9", "Myeloid_c2_LAM2_APOE", "Myeloid_c12_Monocyte_1_IL1B") ~ "Monocytes",
    celltype_subset %in% c("Myeloid_c3_cDC1_CLEC9A") ~ "cDC1",
    celltype_subset %in% c("Myeloid_c11_cDC2_CD1C") ~ "cDC2",
    celltype_subset %in% c("Myeloid_c4_DCs_pDC_IRF7") ~ "pDC",
    celltype_subset %in% c("Myeloid_c0_DC_LAMP3") ~ "mregDC",
    .default = celltype_major
  )
)
```

```{r}
#Subset
DefaultAssay(seurat.BC) <- "RNA"
CD4 <- seurat.BC$celltype_minor == "T cells CD4+"
seurat.BC <- seurat.BC[, CD4]

#Integrate, scale and Findvariablegenes
seurat.BC <- SplitObject(seurat.BC, split.by = "Patient")
seurat.BC <- lapply(seurat.BC, SCTransform)

#Deleted all sample with less than 100 CD4 cells
seurat.BC$CID4523 <- NULL
seurat.BC$CID3946 <- NULL
seurat.BC$CID4461 <- NULL
seurat.BC$CID4465 <- NULL

#Integrate data
features  <- SelectIntegrationFeatures(seurat.BC, nfeatures = 3000)
seurat.BC <- PrepSCTIntegration(seurat.BC, anchor.features = features)
anchors <- FindIntegrationAnchors(seurat.BC, normalization.method = "SCT", anchor.features = features)
seurat.BC.combined <- IntegrateData(anchorset = anchors, normalization.method = "SCT", k.weight = 46)

#Run PCA
seurat.BC.combined <- RunPCA(seurat.BC.combined)

pcas <- 20
ElbowPlot(seurat.BC.combined, ndims = 50) + theme_linedraw() +
  geom_vline(xintercept = pcas, color = "red", size = 0.5) +
  ggtitle("Elbow plot of first 50 PCs")

#Represent UMAP
 seurat.BC.combined <- RunUMAP(
   seurat.BC.combined, dims = seq(pcas), verbose = TRUE
 )
  DimPlot(
  seurat.BC.combined, group.by = "Patient", reduction = "umap", cols = color.list(), 
  raster = TRUE
)
  
  DimPlot(
  seurat.BC.combined, group.by = "celltype_subset", reduction = "umap", cols = color.list(), 
  raster = TRUE
)  
  
FeaturePlot(
  object = seurat.BC.combined, 
  reduction = "umap",
  features = "CXCR6",
  pt.size = 1,
  min.cutoff = 0,
  cols = c("lightgrey", "peru", "orangered4"), 
  order = T
  )

FeaturePlot(
  object = seurat.BC.combined, 
  reduction = "umap",
  features = "CCR7",
  pt.size = 0.35,
  min.cutoff = 0,
  cols = c("lightgrey", "peru", "orangered4"), 
  order = T
  )
```

##Find clusters
```{r}
ress <- c( 1)
kList <- c(80)
ress <- c(0.8, 1, 1.2, 1.5)
kList <- c(20, 50, 80)
for (k in kList) {
  seurat.BC.combined <- FindNeighbors(
    seurat.BC.combined, 
    reduction = "pca", 
    dims = 1:pcas, 
    # features = useGenes, 
    compute.SNN = TRUE, 
    k.param = k, 
    # return.neighbor = TRUE,
    nn.method = "annoy", 
    n.trees = 80,
    graph.name = paste("k", k, sep = "."),
    # do.plot = T
  )  
  cat(paste(">>> With k =", k, "\n"))
  for (res in ress) {
    if (res %in% colnames(seurat.BC.combined@meta.data)) {
      cat("Resolution already done\n")
      cat(
        paste(
          "res:", res, "Number of clusters detected:", 
          length(
            levels(factor(seurat.BC.combined@meta.data[, paste("res", res, sep=".")]))
          ), "\n"
        )
      )
    } else {
      seurat.BC.combined <- FindClusters(
        seurat.BC.combined, 
        temp.file.location = objectsPath, 
        algorithm = 2, 
        n.iter = 20,
        resolution = res, 
        verbose = FALSE, 
        random.seed = 123456,
        graph.name = paste("k", k, sep = ".")
      )
      cat(
        paste(
          "   res:", res, "Number of clusters detected:", 
          length(
            levels(factor(seurat.BC.combined@meta.data[, paste0("k.", k, "_res.", res)]))
          ), "\n"
        )
      )
    }
  }
}

#Plot clusters: 80 0.8 and 80 0.5 
for (k in kList) {
  for (res in ress) {
    p3 <- DimPlot(
      seurat.BC.combined, 
      group.by = paste0("k.", k, "_res.", res), 
      reduction = "umap",
      cols = color.list()
    ) + ggtitle(paste0("umap(", paste0("k.", k, "_res.", res), ")"))
    print(ggarrange(plotlist = list(p3)))
  }
}
```

##Define resolution

```{r}
#Define new cluster as res.Final.1
seurat.BC.combined$res.Final.1 <- factor(
  paste0("C", seurat.BC.combined$k.50_res.1.5), 
  levels = paste0("C", seq(0,  length(unique(seurat.BC.combined$k.50_res.1.5))))
)

Idents(seurat.BC.combined) <- seurat.BC.combined$res.Final.1

#Define original clusters + new cluster C16 as res.Final.2
seurat.BC.combined$res.Final.2 <- ifelse(seurat.BC.combined$res.Final.1 == "C16", "CD4+ Trm", seurat.BC.combined$celltype_subset )

#Order clusters for plotting
seurat.BC.combined@active.ident <- factor(seurat.BC.combined@active.ident, 
                            levels=c('CD4+ Trm',
                                  'T_cells_c3_CD4+_Tfh_CXCL13',
                                  'T_cells_c0_CD4+_CCR7',
                                  'T_cells_c1_CD4+_IL7R',
                                  'T_cells_c2_CD4+_T-regs_FOXP3'))

```

##Define colors

```{r}
colorsPlots <- list(
  Original_clusters = c(
    "T_cells_c1_CD4+_IL7R" = color.list()[3],
    "T_cells_c0_CD4+_CCR7" = color.list()[12],
    "T_cells_c2_CD4+_T-regs_FOXP3" = color.list()[8], 
    "T_cells_c3_CD4+_Tfh_CXCL13" = color.list()[13]
  ),
  New_clusters = c(
    "CD4+ Trm" = color.list()[6],
    "T_cells_c1_CD4+_IL7R" = color.list()[3],
    "T_cells_c0_CD4+_CCR7" = color.list()[12],
    "T_cells_c2_CD4+_T-regs_FOXP3" = color.list()[8], 
    "T_cells_c3_CD4+_Tfh_CXCL13" = color.list()[13]
  ), 
  Sub_clusters = c(
    "C0" = color.list()[2],
    "C1" = color.list()[3],
    "C2" = color.list()[4],
    "C3" = color.list()[5],
    "C4" = color.list()[1],
    "C5" = color.list()[7],
    "C6" = color.list()[8],
    "C7" = color.list()[9],
    "C8" = color.list()[10],
    "C9" = color.list()[11],
    "C10" = color.list()[12],
    "C11" = color.list()[13],
    "C12" = color.list()[14],
    "C13" = color.list()[15],
    "C14" = color.list()[16],
    "C15" = color.list()[17],
    "C16" = color.list()[6],
    "C17" = color.list()[18]
  )
)
```

##Make plots

```{r}
DimPlot(
  seurat.BC.combined, reduction = "umap", group.by = "celltype_subset", 
  cols = colorsPlots$Original_clusters
) + ggtitle("Original identification")

DimPlot(
  seurat.BC.combined, reduction = "umap", group.by = "res.Final.1", 
  cols = colorsPlots$Sub_clusters, pt.size = 1
) + ggtitle("New identification")
```

## CD4 Trm (Human from Savas et al and cluster 0 from our data) of each cluster of CD4 T cells

```{r}
#Checked here how can one avoid the missing genes errors https://github.com/satijalab/seurat/issues/4881
seurat.BC.combined <- AddModuleScore(
  seurat.BC.combined, features = geneSignaturesF, name = "modules",
  ctrl = 50, assay = "SCT"
)
tmp <- tail(seq_len(ncol(seurat.BC.combined@meta.data)), length(geneSignaturesF))
newColumns <- seurat.BC.combined@meta.data[, tmp]
colnames(newColumns) <- names(geneSignaturesF)
seurat.BC.combined@meta.data <- cbind(seurat.BC.combined@meta.data, newColumns)
seurat.BC.combined@meta.data[, tmp] <- NULL
```

## Heatmap modulescores on all subclusters
```{r, fig.height=15, fig.width=20}
#In order to represent the values of module score per each of the sublcusters, we will represent the average in a heatmap. 
scoreexp <- data.frame(seurat.BC.combined@meta.data$res.Final.1)
scoreexp <- cbind("seurat_cluster" = scoreexp, "Mouse_CD4_Trm" = seurat.BC.combined$Costum_C0_up, "Human_CD4_Trm"= seurat.BC.combined$CD4_Trm_Signature)
names(scoreexp)[names(scoreexp) == 'seurat.BC.combined.meta.data.res.Final.1'] <- 'seurat_cluster'

scoreexp <-scoreexp %>% group_by(seurat_cluster) %>% summarise_all(mean)
scoreexp <- as.data.frame(scoreexp)
row.names(scoreexp) <- scoreexp$seurat_cluster
scoreexp$seurat_cluster <- NULL
scoreexp <- as.matrix(scoreexp)
scoreexp <- t(scoreexp)

col_fun = colorRamp2(c(-2, 0, 2.5), c("#5c8f9c", "white", "#654321"))

#We scale the matrix to plot Z-scores
"lightgrey", "peru", "orangered4"
scaled_score = t(scale(t(scoreexp)))

#PLot
ComplexHeatmap::Heatmap(
  scaled_score, 
  col = col_fun,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10), name = "Scaled mean module score",
  column_title_gp = gpar(fontface = "bold", fontsize = 15), 
  column_title = "Expression of CD4Trm signatures",
  column_order = NULL,
  row_order = NULL,
  column_names_rot = 45,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  row_dend_reorder = FALSE,
  row_title = NULL,
  row_names_side = "left",
  border = TRUE,
  width = unit(7, "cm"), height = unit(1, "cm"),
  heatmap_legend_param = list(
    legend_height = unit(2, "cm"),
    title_gp = gpar(fontsize = 10)
  )
)

dev.off()
```

## Correlation cDC1 signature 

### Find frequency of new clusters in every patient
```{r, fig.width=6, fig.height=3.5}
dfProp_c <- seurat.BC.combined@meta.data %>%
  group_by(Patient, res.Final.2) %>%
  summarise(n = n()) %>%
  mutate(freq = (n / sum(n) * 100)) %>%
  ungroup() %>%
  complete(Patient, res.Final.2, fill = list(n = 0, freq = 0)) 

dfProp_c <- as.data.frame(dfProp_c)
```

### Get frequency of each predefined cluster per patient. 
```{r, fig.width=6, fig.height=3.5}
#Get frequencies from defined clusters within TCs. 
#Workaround to get those clusters that have 0 cells in the specific patient inspired from https://kieranhealy.org/blog/archives/2018/11/19/zero-counts-in-dplyr/

#CD4 T cell freq in tumor
dfProp_min <- as.data.frame(dfProp_min)
Tcprop <- dfProp_min[dfProp_min$celltype_minor == "T cells CD4+",]
rownames(Tcprop) <- Tcprop$Patient 
Tcprop$Patient <- NULL
Tcprop$celltype_minor <- NULL
Tcprop$n <- NULL
names(Tcprop) <- "CD4"
CD4_001 <- Tcprop

#cDC1 prop
dfProp_sub <- as.data.frame(dfProp_sub)
cDC1prop <- dfProp_sub[dfProp_sub$celltype_subset == "Myeloid_c3_cDC1_CLEC9A",]
rownames(cDC1prop) <- cDC1prop$Patient 
cDC1prop$Patient <- NULL
cDC1prop$celltype_subset <- NULL
cDC1prop$n <- NULL
names(cDC1prop) <- "cDC1prop"

#cDC2 prop
dfProp_sub <- as.data.frame(dfProp_sub)
cDC2prop <- dfProp_sub[dfProp_sub$celltype_subset == "Myeloid_c11_cDC2_CD1C",]
rownames(cDC2prop) <- cDC2prop$Patient 
cDC2prop$Patient <- NULL
cDC2prop$celltype_subset <- NULL
cDC2prop$n <- NULL
names(cDC2prop) <- "cDC2prop"


#Normalize to obtain % of each res.Final.2 cluster within tumor
dfProp_c$n <- NULL
dfProp_c_wide <- pivot_wider(dfProp_c, names_from = "res.Final.2", values_from = "freq")
dfProp_c_wide <- as.data.frame(dfProp_c_wide)
rownames(dfProp_c_wide) <- dfProp_c_wide$Patient
dfProp_c_wide$Patient <- NULL

cluster_intumor <- merge(dfProp_c_wide, CD4_001, by=0)
rownames(cluster_intumor) <- cluster_intumor$Row.names
cluster_intumor$`CD4+ Trm` <- cluster_intumor$`CD4+ Trm` * cluster_intumor$CD4* 0.01
cluster_intumor$`T_cells_c3_CD4+_Tfh_CXCL13` <- cluster_intumor$`T_cells_c3_CD4+_Tfh_CXCL13` * cluster_intumor$CD4* 0.01
cluster_intumor$`T_cells_c0_CD4+_CCR7` <- cluster_intumor$`T_cells_c0_CD4+_CCR7` * cluster_intumor$CD4* 0.01
cluster_intumor$`T_cells_c1_CD4+_IL7R` <- cluster_intumor$`T_cells_c1_CD4+_IL7R` * cluster_intumor$CD4* 0.01
cluster_intumor$`T_cells_c2_CD4+_T-regs_FOXP3` <- cluster_intumor$`T_cells_c2_CD4+_T-regs_FOXP3` * cluster_intumor$CD4* 0.01

rownames(cluster_intumor) <- cluster_intumor$Row.names 
cluster_intumor$Row.names <- NULL

cluster_intumor <- merge(cluster_intumor, cDC1prop, by=0 )
rownames(cluster_intumor) <- cluster_intumor$Row.names 
cluster_intumor$Row.names <- NULL

cluster_intumor <- merge(cluster_intumor, cDC2prop, by=0 )
rownames(cluster_intumor) <- cluster_intumor$Row.names 
cluster_intumor$Row.names <- NULL
```


### Merge and find correlation of scRNAseq
```{r, fig.width=6, fig.height=3.5}
# Create dataframe where to save results
res <- data.frame(matrix(nrow = 0, ncol = 4))
colnames(res) <- c("var1", "var2", "correlation", "pvalue")

# Correlation in loop
for(i in colnames(cluster_intumor[,1:6])) {
  for(j in c("cDC1prop", "cDC2prop")) {
    a <- cor.test(cluster_intumor[[i]], cluster_intumor[[j]], method = "spearman")
    res <- rbind(res, data.frame(
      "var1" = i,
      "var2" = j,
      "correlation" = a$estimate,
      "pvalue" = a$p.value))
  }
}

# Remove rownames
rownames(res) <- NULL

res <- cbind(res,log10(res$pvalue))
res<- cbind(res, res$`log10(res$pvalue)`*-1)
colnames(res)[5]<- "mistake"
colnames(res)[6]<- "-log10padj.value"
```


### Visualization of correlation scRNAseq
```{r fig.height=13}
# Visual representation of correlation
p1<- ggscatter(
  cluster_intumor, x= "cDC1prop", 
  y = "CD4+ Trm",
  color = "black", 
  size = 1,
  #rug = TRUE,
  #fill = "Tissue",
  add = "reg.line", conf.int = TRUE, 
  #palette = color.list(),
  cor.coef = TRUE, cor.method = "spearman", 
  title = "Correlation cDC1"
) + #geom_smooth(method = "lm", colour = mycol, se = FALSE) + 
  theme_classic2() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```



