---
title: "Heras-Murillo et al. Mouse scRNAseq"
author: "Ignacio Heras-Murillo"
documentclass: article
output:
  html_document: 
    number_sections: yes
    self_contained: no
    theme: united
    toc: yes
    toc_depth: 5  
    toc_float: true
    collapsed: no
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 4
urlcolor: blue
classoption: a4paper
---

#Load packages
```{r setup, include=FALSE}
## packages
suppressMessages(library("Matrix"))
suppressMessages(library("ggplot2"))
suppressMessages(library("ggrepel"))
suppressMessages(library("ggpubr"))
suppressMessages(library("ggridges"))
suppressMessages(library("dplyr"))
suppressMessages(library("tidyr"))
suppressMessages(library("readr"))
suppressMessages(library("biomaRt"))
suppressMessages(library("Seurat"))
suppressMessages(library("scater"))
suppressMessages(library("org.Mm.eg.db"))
suppressMessages(library("ggvenn"))
suppressMessages(library("gtools"))
suppressMessages(library("viridis"))
suppressMessages(library("plotly"))
suppressMessages(library("Matrix.utils"))
suppressMessages(library("tibble"))
suppressMessages(library("ggpointdensity"))
suppressMessages(library("melt"))
suppressMessages(library("reshape2"))
suppressMessages(library("circlize"))
suppressMessages(library("KEGGREST"))
suppressMessages(library("EnrichmentBrowser"))
suppressMessages(library("fgsea"))
suppressMessages(library("reshape2"))

## paths
projectPath <- here::here()
dataPath <- here::here("data")
dataChip1Path <- here::here("data/Chip1-Data")
dataChip2Path <- here::here("data/Chip2-Data")
metadataPath <- here::here("metadata")
sourceCode <- here::here("src")
objectsPath <- here::here("HW-Fig6")
processedPath <- here::here("HW-Fig6/processed_data")
reportsPath <- here::here("reports/HW-Fig6")
plotsPath <- here::here(reportsPath, "plots")
prefix <- "HW-Fig6"

## scr code
source(file.path(sourceCode, "helperFunctions.R"))

## rmarkdown settings
knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.path = file.path(reportsPath, "plots/"),
  fig.pos = "H",
  message = FALSE,
  warning = FALSE,
  comment = NA,
  dev = "png",
  dpi = 300
)

## global options
options(encoding = 'UTF-8')
```

```{r}
## just in case
if (!file.exists(reportsPath)) {
  dir.create(reportsPath)
  dir.create(file.path(reportsPath, "plots"))
}
```

```{r}
if (file.exists(file.path(
  objectsPath, paste(prefix, "bmAnnotations.rds", sep = ".")
))) {
  message(paste("Load bmAnnotations from file:", file.path(
    objectsPath, paste(prefix, "bmAnnotations.rds", sep = ".")
  )))
  bmAnnotations <- readRDS(file = file.path(
    objectsPath, paste(prefix, "bmAnnotations.rds", sep = ".")
  ))
  MTGenes <- readRDS(file = file.path(
    objectsPath, paste(prefix, "mtGenes.rds", sep = ".")
  ))
  RBGenes1 <- readRDS(file = file.path(
    objectsPath, paste(prefix, "rbGenes.1.rds", sep = ".")
  ))
  RBGenes2 <- readRDS(file = file.path(
    objectsPath, paste(prefix, "rbGenes.2.rds", sep = ".")
  ))
} else {
  ensembl <- useEnsembl(biomart = "genes")
  ensembl <- useDataset(dataset = "mmusculus_gene_ensembl", mart = ensembl)
  
  bmAnnotations <- getBM(
    attributes = c(
      "external_gene_name", "entrezgene_id",
      "entrezgene_trans_name", "entrezgene_accession",
      "ensembl_gene_id", "gene_biotype", 
      "chromosome_name", "start_position", 
      "end_position", "strand", 
      "description"
    ), uniqueRows = T, mart = ensembl
  )
  # bmAnnotations <- bmAnnotations %>% filter(!grepl("^CHR_", chromosome_name))
  bmAnnotations$external_gene_name <- toupper(bmAnnotations$external_gene_name)
  ## Mitochondrial genes
  MTGenes <- bmAnnotations[bmAnnotations$chromosome_name == "MT",]
  RBGenes1 <- bmAnnotations %>% filter(gene_biotype == "rRNA")
  RBGenes2 <- read.delim(
    file.path(metadataPath, "AllRibosomalProteins.csv"),
    sep = "\t", stringsAsFactors = F, header = T
  )
  saveRDS(MTGenes, file = file.path(
    objectsPath, paste(prefix, "mtGenes.rds", sep = ".")
  ))
  saveRDS(RBGenes1, file = file.path(
    objectsPath, paste(prefix, "rbGenes.1.rds", sep = ".")
  ))
  saveRDS(RBGenes2, file = file.path(
    objectsPath, paste(prefix, "rbGenes.2.rds", sep = ".")
  ))
  saveRDS(bmAnnotations, file = file.path(
    objectsPath, paste(prefix, "bmAnnotations.rds", sep = ".")
  ))
}
```

# Load data

```{r}
seuratSC.Raw.1 <- readRDS(
  file = file.path(objectsPath, paste(prefix, "seuratSC.CMO.Norm.1.rds", sep = "."))
)
seuratSC.Raw.2 <- readRDS(
  file = file.path(objectsPath, paste(prefix, "seuratSC.CMO.Norm.2.rds", sep = "."))
)
```

Merge reads from both sequencing lines.
```{r}
seuratSC.Raw.comb <- merge(
  x = seuratSC.Raw.1, y = seuratSC.Raw.2, 
  add.cell.ids = NULL, project = "TILs-CancerVacc"
)

rm(seuratSC.Raw.1)
rm(seuratSC.Raw.2)
```

# Identify cells per condition and quality control

Identify the CMO codes per experimental condition and biological replicate.
```{r}
idCMOs <- read.csv(
  file = file.path(metadataPath, "identification_cmos.csv"), header = TRUE
)
```

```{r}
seuratSC.Raw.comb@meta.data$Condition <- "Unknown"
for (i in seq(nrow(idCMOs))) {
  mod <- seuratSC.Raw.comb@meta.data$hash.ID == idCMOs[i, "cmo_ids"]
  seuratSC.Raw.comb@meta.data[mod, "Condition"] <- idCMOs[i, "sample_id"]
}
```

Remove cells that do not have a unique CMO

```{r}
validCells <- seuratSC.Raw.comb@meta.data %>% filter(
  !Condition %in% "Unknown"
) %>% pull(Cell_ID)
```

```{r}
seuratSC.Raw.comb.F <- seuratSC.Raw.comb[, validCells]

rm(seuratSC.Raw.comb)
```

Create the column "TypeVacc", which refers to the experimental group each cell belongs to. 

```{r}
seuratSC.Raw.comb.F@meta.data$TypeVacc <- sapply(
  X = strsplit(x = seuratSC.Raw.comb.F@meta.data$Condition, split = "_"), 
  FUN = function(x) x[[1]]
)
```

## Create Seurat object and pre-processing

```{r}
genesMetadata <- readRDS(
  file = file.path(objectsPath, paste0(prefix, ".genesMetadata.rds"))
)
```

Create seurat object

```{r}
genesMetadataF <- genesMetadata[match(rownames(seuratSC.Raw.comb.F), genesMetadata$ENSEMBL), ]
rownames(genesMetadataF) <- genesMetadataF$ENSEMBL
genesMetadataFNAs <- genesMetadataF[!is.na(genesMetadataF$SYMBOL),]
seuratSC.Raw.comb.F.Genes <- seuratSC.Raw.comb.F[genesMetadataFNAs$ENSEMBL, ]
rm(seuratSC.Raw.comb.F)
```

```{r}
rawCountsCheck <- seuratSC.Raw.comb.F.Genes@assays$RNA@counts
rownames(rawCountsCheck) <- genesMetadataFNAs$SYMBOL
aggNames <- factor(genesMetadataFNAs$SYMBOL)
rawCountsAgg <- Matrix.utils::aggregate.Matrix(
  x = rawCountsCheck, groupings = aggNames, fun = "sum"
)
```

```{r}
seuratSC.Raw.comb.Final <- CreateSeuratObject(
  counts = rawCountsAgg, 
  meta.data = seuratSC.Raw.comb.F.Genes@meta.data, 
  min.features = 1, 
  min.cells = 1,
  names.field = 1, 
  names.delim = "-",
  project = "Vaccines_001"
)
seuratSC.Raw.comb.Final

rm(seuratSC.Raw.comb.F.Genes)

```
## Normalization and scaling

```{r}
seuratSC.Norm.comb <- NormalizeData(
  object = seuratSC.Raw.comb.Final,
  normalization.method = "LogNormalize",
  vscale.factor = 10000,
  verbose = TRUE
)

rm(seuratSC.Raw.comb.Final)
```

## Find variable features
```{r}
seuratSC.Norm.comb <- FindVariableFeatures(
  seuratSC.Norm.comb, selection.method = "vst"
)
```
## Scale data

```{r}
seuratSC.Scal.comb <- ScaleData(
    object = seuratSC.Norm.comb, 
    do.scale = TRUE,
    do.center = TRUE,
    display.progress = TRUE,
    vars.to.regress = "nCount_RNA",
  )
rm(seuratSC.Norm.comb)
```

## PCA
```{r}
seuratSC.Scal.comb <- RunPCA(
  seuratSC.Scal.comb,
  ndims.print = 1:5,
  nfeatures.print = 30,
  npcs = 50
  # reduction.name = "pca_varGenes",
  # reduction.key = paste("PC_varGenes", i, sep = "_")
)
```

# Clustering and 2D representation

## Find clusters
```{r SelectClusteringResolution}
pcas <- 1:25
ress <- c(0.5)
kList <- c(15)

for (k in kList) {
  seuratSC.Scal.comb <- FindNeighbors(
    seuratSC.Scal.comb, 
    reduction = "pca", 
    dims = pcas, 
    # features = useGenes, 
    compute.SNN = TRUE, 
    k.param = k, 
    # return.neighbor = TRUE,
    nn.method = "annoy", 
    n.trees = 80,
    graph.name = paste("k", k, sep = "."),
    # do.plot = T
  )  
  cat(paste(">>> With k =", k, "\n"))
  for (res in ress) {
    if (res %in% colnames(seuratSC.Scal.comb@meta.data)) {
      cat("Resolution already done\n")
      cat(
        paste(
          "res:", res, "Number of clusters detected:", 
          length(
            levels(factor(seuratSC.Scal.comb@meta.data[, paste("res", res, sep=".")]))
          ), "\n"
        )
      )
    } else {
      seuratSC.Scal.comb <- FindClusters(
        seuratSC.Scal.comb, 
        temp.file.location = objectsPath, 
        algorithm = 2, 
        n.iter = 20,
        resolution = res, 
        verbose = FALSE, 
        random.seed = 123456,
        graph.name = paste("k", k, sep = ".")
      )
      cat(
        paste(
          "   res:", res, "Number of clusters detected:", 
          length(
            levels(factor(seuratSC.Scal.comb@meta.data[, paste0("k.", k, "_res.", res)]))
          ), "\n"
        )
      )
    }
  }
}
```

## UMAP
```{r}
seed <- 9876
n.neighList <- c(15L)
m.distList <- c(0.6)
  for (n in n.neighList) {
    for (dist in m.distList) {
         seuratSC.Scal.comb <- RunUMAP(
           seuratSC.Scal.comb, 
           seed.use = seed, 
           dims = pcas, 
           n.neighbors = n,
           min.dist = dist,
           n.epochs = 400,
           reduction.name = paste("umap", n, dist, sep = ".")
         )
      }  
    }
```

Set final resolution
```{r}
seuratSC.Scal.comb$res.Final.1 <- factor(
  paste0("C", seuratSC.Scal.comb$k.15_res.0.5), 
  levels = paste0("C", seq(0,  length(unique(seuratSC.Scal.comb$k.15_res.0.5))))
)

Idents(seuratSC.Scal.comb) <- seuratSC.Scal.comb$res.Final.1
seuratSC.Scal.comb@reductions$umap.Final <- seuratSC.Scal.comb@reductions$umap.15.0.6
```

Plot results
```{r}
DimPlot(
  seuratSC.Scal.comb, reduction = "umap.Final", label = TRUE, cols = color.list(), label.box = TRUE
)
```

#Identification of cells

##FindAllMarkers
```{r}
markersGeneral <- FindAllMarkers(
    seuratSC.Scal.comb, only.pos = FALSE, 
    min.pct = 0.25, logfc.threshold = 0.25, test.use = "wilcox"
  )
```

## Plot the most significant upregulated genes in each cluster
```{r}
markersSignificant <- markersGeneral %>% 
  filter(abs(avg_log2FC) >= 1, p_val_adj <= 0.05) %>% 
  arrange(desc(avg_log2FC))

plotGenes <- markersSignificant %>% group_by(cluster) %>%
    slice_max(n = 4, order_by = avg_log2FC) %>% pull(gene)
```

```{r, fig.width=10, fig.height=10, warning=FALSE, error=FALSE}
DotPlotDef(seuratSC.Scal.comb, features = unique(plotGenes)) +
  theme(axis.text.x = element_text(size = 9), axis.text.y = element_text(size = 9)) + 
  scale_color_gradientn(
    colors = colorRampPalette(
      colors = c("#1189d4", "#1cad3e", "yellow", "orange", "red")
      )(50)) + 
  ggtitle("Top markers per cluster") + coord_flip()

#ggsave("Data_cluster_genes.png", path = plotsPath, width=10, height=10)

```

## (!) Identification of contaminating clusters

We found a major set of clusters (C8, C13 and C14) and two extra clusters (C9, C10 and C12) that lack the expression of CD3 and either CD4 or CD8. Therefore, we considered them contaminations and were removed from the analysis. 
```{r}
FeaturePlotDef(obj = seuratSC.Scal.comb, feature = "CD3E", dim.reduction = "umap.Final")
FeaturePlotDef(obj = seuratSC.Scal.comb, feature = "CD4", dim.reduction = "umap.Final") 
FeaturePlotDef(obj = seuratSC.Scal.comb, feature = "CD8A", dim.reduction = "umap.Final")
```

## Removal of contaminating cells and re-analysis

###  Identification of contaminating cells
```{r}
seuratSC.Scal.comb@meta.data <- mutate(
  seuratSC.Scal.comb@meta.data,
  Cell_type_Trim = case_when(
    res.Final.1 %in% c("C12", "C9", "C10", "C8", "C13", "C14") ~ "Delete",
    res.Final.1 %in% c("C0","C1", "C2", "C3", "C4","C5", "C6", "C7", "C11") ~ "Keep",
    TRUE ~ NA_character_
  )
)

ValidCellsMarkers <- seuratSC.Scal.comb@meta.data %>% filter (
  !Cell_type_Trim %in% "Delete"
) %>% pull(Cell_ID)

rm(seuratSC.Scal.comb)
```

### Re-load raw data

```{r}
seuratSC.Raw.1 <- readRDS(
  file = file.path(objectsPath, paste(prefix, "seuratSC.CMO.Norm.1.rds", sep = "."))
)
seuratSC.Raw.2 <- readRDS(
  file = file.path(objectsPath, paste(prefix, "seuratSC.CMO.Norm.2.rds", sep = "."))
)
```

Merge reads from both sequencing lines.
```{r}
seuratSC.Raw.comb <- merge(
  x = seuratSC.Raw.1, y = seuratSC.Raw.2, 
  add.cell.ids = NULL, project = "TILs-CancerVacc"
)

rm(seuratSC.Raw.1)
rm(seuratSC.Raw.2)
```

### Re-Identify cells per condition and quality control

Identify the CMO codes per experimental condition and biological replicate.
```{r}
idCMOs <- read.csv(
  file = file.path(metadataPath, "identification_cmos.csv"), header = TRUE
)
```

```{r}
seuratSC.Raw.comb@meta.data$Condition <- "Unknown"
for (i in seq(nrow(idCMOs))) {
  mod <- seuratSC.Raw.comb@meta.data$hash.ID == idCMOs[i, "cmo_ids"]
  seuratSC.Raw.comb@meta.data[mod, "Condition"] <- idCMOs[i, "sample_id"]
}
```

Remove cells that do not have a unique CMO and that are not contaminating cells (not expressing CD3 and CD4 or CD8)

```{r}
#seuratSC.Raw.comb.F <- seuratSC.Raw.comb[, validCells]
seuratSC.Raw.comb.F <- seuratSC.Raw.comb[, ValidCellsMarkers]

rm(seuratSC.Raw.comb)
```

Create the column "TypeVacc", which refers to the experimental group each cell belongs to. 

```{r}
seuratSC.Raw.comb.F@meta.data$TypeVacc <- sapply(
  X = strsplit(x = seuratSC.Raw.comb.F@meta.data$Condition, split = "_"), 
  FUN = function(x) x[[1]]
)
```

### Create Seurat object and pre-processing

```{r}
genesMetadata <- readRDS(
  file = file.path(objectsPath, paste0(prefix, ".genesMetadata.rds"))
)
```

Create seurat object

```{r}
genesMetadataF <- genesMetadata[match(rownames(seuratSC.Raw.comb.F), genesMetadata$ENSEMBL), ]
rownames(genesMetadataF) <- genesMetadataF$ENSEMBL
genesMetadataFNAs <- genesMetadataF[!is.na(genesMetadataF$SYMBOL),]
seuratSC.Raw.comb.F.Genes <- seuratSC.Raw.comb.F[genesMetadataFNAs$ENSEMBL, ]
rm(seuratSC.Raw.comb.F)
```

```{r}
rawCountsCheck <- seuratSC.Raw.comb.F.Genes@assays$RNA@counts
rownames(rawCountsCheck) <- genesMetadataFNAs$SYMBOL
aggNames <- factor(genesMetadataFNAs$SYMBOL)
rawCountsAgg <- Matrix.utils::aggregate.Matrix(
  x = rawCountsCheck, groupings = aggNames, fun = "sum"
)
```

```{r}
seuratSC.Raw.comb.Final <- CreateSeuratObject(
  counts = rawCountsAgg, 
  meta.data = seuratSC.Raw.comb.F.Genes@meta.data, 
  min.features = 1, 
  min.cells = 1,
  names.field = 1, 
  names.delim = "-",
  project = "Vaccines_001"
)
seuratSC.Raw.comb.Final

rm(seuratSC.Raw.comb.F.Genes)


#saveRDS(seuratSC.Raw.comb.Final, file = file.path(objectsPath, paste(prefix, "seuratSC.Raw.comb.Final.rds", sep = ".")))
```

### Normalization and scaling

```{r}
seuratSC.Norm.comb <- NormalizeData(
  object = seuratSC.Raw.comb.Final,
  normalization.method = "LogNormalize",
  vscale.factor = 10000,
  verbose = TRUE
)

rm(seuratSC.Raw.comb.Final)
```

### Find variable features
```{r}
seuratSC.Norm.comb <- FindVariableFeatures(
  seuratSC.Norm.comb, selection.method = "vst"
)
```
### Scale data

```{r}
seuratSC.Scal.comb <- ScaleData(
    object = seuratSC.Norm.comb, 
    do.scale = TRUE,
    do.center = TRUE,
    display.progress = TRUE,
    vars.to.regress = "nCount_RNA",
  )
rm(seuratSC.Norm.comb)
```
### PCA

```{r}
seuratSC.Scal.comb <- RunPCA(
  seuratSC.Scal.comb,
  ndims.print = 1:5,
  nfeatures.print = 30,
  npcs = 50
  # reduction.name = "pca_varGenes",
  # reduction.key = paste("PC_varGenes", i, sep = "_")
)
```

### Clustering and 2D representation

### Find clusters
```{r SelectClusteringResolution}
pcas <- 1:25
ress <- c(1)
kList <- c(50)

for (k in kList) {
  seuratSC.Scal.comb <- FindNeighbors(
    seuratSC.Scal.comb, 
    reduction = "pca", 
    dims = pcas, 
    # features = useGenes, 
    compute.SNN = TRUE, 
    k.param = k, 
    # return.neighbor = TRUE,
    nn.method = "annoy", 
    n.trees = 80,
    graph.name = paste("k", k, sep = "."),
    # do.plot = T
  )  
  cat(paste(">>> With k =", k, "\n"))
  for (res in ress) {
    if (res %in% colnames(seuratSC.Scal.comb@meta.data)) {
      cat("Resolution already done\n")
      cat(
        paste(
          "res:", res, "Number of clusters detected:", 
          length(
            levels(factor(seuratSC.Scal.comb@meta.data[, paste("res", res, sep=".")]))
          ), "\n"
        )
      )
    } else {
      seuratSC.Scal.comb <- FindClusters(
        seuratSC.Scal.comb, 
        temp.file.location = objectsPath, 
        algorithm = 2, 
        n.iter = 20,
        resolution = res, 
        verbose = FALSE, 
        random.seed = 123456,
        graph.name = paste("k", k, sep = ".")
      )
      cat(
        paste(
          "   res:", res, "Number of clusters detected:", 
          length(
            levels(factor(seuratSC.Scal.comb@meta.data[, paste0("k.", k, "_res.", res)]))
          ), "\n"
        )
      )
    }
  }
}
```

### tSNE
```{r}
seed <- 9876
perpList <- c(120)
  for (perpl in perpList) {
       seuratSC.Scal.comb <- RunTSNE(
         seuratSC.Scal.comb, 
         seed.use = seed, 
         dims = pcas, 
         theta = 0.5,
         tsne.method = "Rtsne", # "Rtsne", # "FIt-SNE", รง, 
         perplexity = perpl, 
         reduction.name = paste("tSNE", perpl, sep=".")
          # , fast_tsne_path = "/opt/FIt-SNE/bin/fast_tsne"
       )
    }
```


### Set final resolution
```{r}
if (file.exists(file.path(objectsPath, paste0(prefix, ".final.seuratSC.Scal.comb.rds")))) {
  seuratSC.Scal.comb <- readRDS(file.path(objectsPath, paste0(prefix, ".final.seuratSC.Scal.comb.rds")))
} else {
  saveRDS(seuratSC.Scal.comb, file = file.path(objectsPath, paste0(prefix, ".final.seuratSC.Scal.comb.rds")))
}

seuratSC.Scal.comb$res.Final.1 <- factor(
  paste0("C", seuratSC.Scal.comb$k.50_res.1), 
  levels = paste0("C", seq(0,  length(unique(seuratSC.Scal.comb$k.50_res.1))))
)

Idents(seuratSC.Scal.comb) <- seuratSC.Scal.comb$res.Final.1
seuratSC.Scal.comb@reductions$tsne.Final <- seuratSC.Scal.comb@reductions$tSNE.120
```

###Identification of cells

###FindAllMarkers
```{r}
markersGeneral <- FindAllMarkers(
    seuratSC.Scal.comb, only.pos = FALSE, 
    min.pct = 0.25, logfc.threshold = 0.25, test.use = "wilcox"
  )
```

# Name clusters 
```{r}
seuratSC.Scal.comb@meta.data <- mutate(
  seuratSC.Scal.comb@meta.data,
  Cell_type_Specific_1 = case_when(
    res.Final.1 == "C0" ~ "CD4+ Trm",
    res.Final.1 == "C1" ~ "Treg",
    res.Final.1 == "C2" ~ "CD8+ Tcm",
    res.Final.1 == "C3" ~ "CD4+ Teff",
    res.Final.1 == "C4" ~ "CD8+ Teff",
    res.Final.1 == "C5" ~ "CD8+ Tcm_KLR",
    res.Final.1 == "C6" ~ "Dividing TC - G2M",
    res.Final.1 == "C7" ~ "CD8+ Trm cytotoxic ",
    res.Final.1 == "C8" ~ "Tc IFN",
    res.Final.1 == "C9" ~ "Dividing TC - G2M CENPA+",
    res.Final.1 == "C10" ~ "Dividing TC - S",
    res.Final.1 == "C11" ~ "CD4+ Teff IL7R",
    res.Final.1 == "C12" ~ "Tc NFKb+",
    TRUE ~ NA_character_
  )
)

#Set colors
Cell_type_Specific_1 = c(
    "CD4+ Trm" = color.list()[6],
    "CD4+ Teff" = color.list()[2],
    "CD4+ Teff IL7R" = color.list()[3],
    "CD8+ Teff" = color.list()[4],
    "CD8+ Trm cytotoxic " = color.list()[5],
    "CD8+ Tcm" = color.list()[1],
    "CD8+ Tcm_KLR" = color.list()[7],
    "Treg" = color.list()[8], 
    "Tc IFN" = color.list()[9],
    "Tc NFKb+" = color.list()[10],
    "Dividing TC - G2M" = color.list()[11],
    "Dividing TC - G2M CENPA+" = color.list()[12],
    "Dividing TC - S" = color.list()[13]
  )
```

```{r}
markersGeneral %>%
    group_by(cluster) %>%
    slice_max(n = 10, order_by = avg_log2FC)
```  

# FIG. S5A - Plot the most significant upregulated genes in each cluster
```{r}
markersSignificant <- markersGeneral %>% 
  filter(abs(avg_log2FC) >= 1, p_val_adj <= 0.05) %>% 
  arrange(desc(avg_log2FC))

seuratSC.Scal.comb@active.ident <- factor(seuratSC.Scal.comb@active.ident, 
                            levels=c('C0','C3','C11','C4','C7','C2','C5','C1','C8','C12', 'C6', 'C9', 'C10'))
markersSignificant$cluster <- factor(markersSignificant$cluster, 
                            levels=c('C0','C3','C11','C4','C7','C2','C5','C1','C8','C12', 'C6', 'C9', 'C10'))

plotGenes <- markersSignificant %>% group_by(cluster) %>% 
    slice_max(n = 4, order_by = avg_log2FC) %>% pull(gene)

```

```{r, fig.width=10, fig.height=10, warning=FALSE, error=FALSE}
DotPlotDef(seuratSC.Scal.comb,     
           clustering.genes = FALSE,
           clustering.cells = FALSE,features = unique(plotGenes)) +
  theme(axis.text.x = element_text(size = 9, angle = 45), axis.text.y = element_text(size = 9)) + 
  scale_color_gradientn(
    colors = colorRampPalette(
      #colors = c("#4E79A7", "#1cad3e", "yellow", "orange", "red")
      colors = c("#5c8f9c", "white", "#654321")
      )(50)) + 
  ggtitle("Top DEG genes per cluster") + coord_flip()

```

# FIG. S5B - Plot the most significant upregulated genes in each cluster
```{r}
markersSignificantC0 <- markersGeneral %>% 
  filter(abs(avg_log2FC) >= 1, p_val_adj <= 0.05) %>% 
  arrange(desc(avg_log2FC)) %>% filter(cluster == "C0")

plotGenesC0 <- markersSignificantC0 %>% pull(gene) %>% {c(head(.,11), tail(.,11))}
```

```{r, fig.width=10, fig.height=10, warning=FALSE, error=FALSE}
DotPlotDef(seuratSC.Scal.comb,     
           clustering.genes = FALSE,
           clustering.cells = FALSE,features = unique(plotGenesC0)) +
  theme(axis.text.x = element_text(size = 9, angle = 45), axis.text.y = element_text(size = 9)) + 
  scale_color_gradientn(
    colors = colorRampPalette(
      #colors = c("#4E79A7", "#1cad3e", "yellow", "orange", "red")
      colors = c("#5c8f9c", "white", "#654321")
      )(50)) + 
  ggtitle("Top markers C0") + coord_flip()
```


# FIG. 6D - Violin plot showing the key markers
```{r, fig.height=15, fig.width=20}
#Inspired from https://github.com/kstromhaug/oliveira-stromhaug-cd4-code/blob/main/plotting/make_voilin_plots_cd4.R

cluster_order = c('CD4+ Trm','CD4+ Teff','CD4+ Teff IL7R','CD8+ Teff','CD8+ Trm cytotoxic ','CD8+ Tcm','CD8+ Tcm_KLR')
vgenes <- c('CD44', 'CD69', 'CXCR6', 'BHLHE40', 'ICOS', 'SELL', 'CCR7', 'S1PR1', 'TCF7', 'KLF2')

### make table for ggplot
setdiff(vgenes, rownames(seuratSC.Scal.comb))
genexp <- data.frame(t(seuratSC.Scal.comb@assays$RNA@data[vgenes, ]))
genexp <- cbind(seuratSC.Scal.comb@meta.data[,'Cell_type_Specific_1'], rownames(genexp), genexp); dim(genexp)
names(genexp)[names(genexp) == 'seuratSC.Scal.comb@meta.data[, \"Cell_type_Specific_1\"]'] <- 'seurat_cluster'
names(genexp)[names(genexp) == 'rownames(genexp)'] <- 'Row.names'
genexp.m <- melt(genexp, id.vars=c('Row.names','seurat_cluster'))
genexp.m$seurat_cluster <- factor(genexp.m$seurat_cluster, levels = cluster_order)
genexp.m <- na.omit(genexp.m)

###Violinplot
ggplot(genexp.m) + geom_violin(aes(x=variable, y=value, fill=seurat_cluster), color='black', scale='width', kernel='gaussian', adjust=1.5) + 
  scale_y_continuous(name="Expression", limits=c(-2, 3), breaks=c(0,3)) +
  theme(axis.ticks = element_blank()) +
  facet_grid(rows='seurat_cluster') + 
  theme_classic() + scale_fill_manual(values=Cell_type_Specific_1) +
  theme(axis.text.x = element_text(angle = 90)) + xlab('Gene') + NoLegend() + 
  theme(panel.spacing.y = unit(0, "lines"), strip.text.y = element_blank())

```

# FIG. S5C - Heatmap general T cell characterization
```{r, fig.height=15, fig.width=20}
#Inspired from https://jokergoo.github.io/ComplexHeatmap-reference/book/a-single-heatmap.html#split-heatmap-annotations and Oliveira et al, 2022
cluster_order = c('CD4+ Trm','CD4+ Teff','CD4+ Teff IL7R','CD8+ Teff','CD8+ Trm cytotoxic ','CD8+ Tcm','CD8+ Tcm_KLR','Treg','Tc IFN','Tc NFKb+', 'Dividing TC - G2M', 'Dividing TC - G2M CENPA+', 'Dividing TC - S')

#General TC markers
vgenes <- c('CD3E','CD4','CD8A','SELL','CCR7','IL7R','CD28','FAS','CD27','ITGAE','ITGAL','ITGAM','ITGAX','PDCD1','TIGIT','HAVCR2', 'LAG3', 'CTLA4', 'KLRG1', 'TNFRSF14', 'BTLA', 'CD160', 'CD38', 'ENTPD1', 'NT5E', 'CD69', 'IL2RA', 'ICOS', 'TNFRSF4', 'TNFRSF9', 'CD40LG', 'CD44', 'GZMA', 'GZMB', 'GZMK', 'GZMM', 'PRF1', 'NKG7', 'IFNG', 'TNF', 'IL17A', 'IL2', 'LEF1', 'TCF7', 'EOMES', 'TBX21', 'PRDM1', 'TOX', 'GATA3', 'ID2', 'ID3', 'NR4A1', 'FOXP3', 'MKI67', 'TOP2A', 'KLRB1', 'KLRC3')

#To check if there is any of the vgenes that is absent in genexp. If so, remove it and proceed. 
setdiff(vgenes, rownames(seuratSC.Scal.comb))

genexp <- data.frame(t(seuratSC.Scal.comb@assays$RNA@data[vgenes, ]))
genexp <- cbind(seuratSC.Scal.comb@meta.data[,'Cell_type_Specific_1'], genexp); dim(genexp)
names(genexp)[names(genexp) == 'seuratSC.Scal.comb@meta.data[, \"Cell_type_Specific_1\"]'] <- 'seurat_cluster'

#Summarize it to get the mean info per cluster and per gene
genexp.h <-genexp %>% group_by(seurat_cluster) %>% summarise_all(mean)
genexp.h <- as.data.frame(genexp.h)
row.names(genexp.h) <- genexp.h$seurat_cluster 
genexp.h$seurat_cluster <- NULL
genexp.h <- as.matrix(genexp.h)
genexp.h <- t(genexp.h)

col_fun = colorRamp2(c(0, 0.5, 1, 2), c("#440154FF", "#31688EFF", "#35B779FF" ,"#FDE725FF"))

#Create object to split columns and rows
column_split <- rep("group1", 13)
column_split[1:3] <- "group2"
column_split[4:7] <- "group3"
column_split[8] <- "group4"
column_split[9:10] <- "group5"
column_split[11:13] <- "group6"


#Row split TC general
row_split <- rep("group1", 57)
row_split[1:3] <- "group2"
row_split[4:13] <- "group3"
row_split[14:22] <- "group4"
row_split[23:32] <- "group5"
row_split[33:41] <- "group6"
row_split[42:55] <- "group7"
row_split[56:57] <- "group8"

#Plot heatmap
ComplexHeatmap::Heatmap(
  genexp.h, 
  col = col_fun,
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10), name = "Mean expression",
  column_title_gp = gpar(fontface = "bold", fontsize = 15), 
  column_title = "Expression of Th signature",
  column_order = cluster_order,
  row_order = vgenes,
  column_names_rot = 45,
  cluster_columns = FALSE,
  show_row_dend = FALSE,
  row_dend_reorder = FALSE,
  column_split = column_split,
  column_gap = unit(2, "mm"), 
  row_split = row_split,
  row_gap = unit(1, "mm"),
  row_title = NULL,
  row_names_side = "left",
  border = TRUE,
  width = unit(6.5, "cm"), height = unit(17, "cm"),
  heatmap_legend_param = list(
  legend_height = unit(2, "cm"),
  title_gp = gpar(fontsize = 10)
  )
)
```

# FIG. 6C - tSNE with cluster annotation
```{r}
DimPlot(
  seuratSC.Scal.comb, reduction = "tsne.Final", group.by = "Cell_type_Specific_1", 
  cols = Cell_type_Specific_1
) + ggtitle("Specific identification")

```

# FIG. 6E - tSNE density plot for each experimental group
```{r, fig.width=6, fig.height=3.5}
#To chose color scales https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
finalPlots <- cbind(pcaSel, tsneSel, umapSel, seuratSC.Scal.comb@meta.data)

ggplot(finalPlots[finalPlots$TypeVacc == "Naive",], aes(x = TSNE_1, y = TSNE_2)) +
  geom_point(aes(x = TSNE_1, y = TSNE_2), data = finalPlots[finalPlots$TypeVacc != "Naive",], color = "lightgrey", size=0.5) + 
  geom_pointdensity(adjust=4, size = 1) +
  scale_color_viridis_c(limits = c(0,325), option = "magma")+
  theme_classic() + ggtitle("TSNE by Naive")

ggplot(finalPlots[finalPlots$TypeVacc == "PBS",], aes(x = TSNE_1, y = TSNE_2)) +
  geom_point(aes(x = TSNE_1, y = TSNE_2), data = finalPlots[finalPlots$TypeVacc != "PBS",], color = "lightgrey", size=0.5) + 
  geom_pointdensity(adjust=4, size = 1) +
  scale_color_viridis_c(limits = c(0,325), option = "magma")+
  theme_classic() + ggtitle("TSNE by PBS")

ggplot(finalPlots[finalPlots$TypeVacc == "PD1",], aes(x = TSNE_1, y = TSNE_2)) +
  geom_point(aes(x = TSNE_1, y = TSNE_2), data = finalPlots[finalPlots$TypeVacc != "PD1",], color = "lightgrey", size=0.5) + 
  geom_pointdensity(adjust=4, size = 1) +
  scale_color_viridis_c(limits = c(0,325), option = "magma")+
  theme_classic() + ggtitle("TSNE by PD1")

ggplot(finalPlots[finalPlots$TypeVacc == "cDC1",], aes(x = TSNE_1, y = TSNE_2)) +
  geom_point(aes(x = TSNE_1, y = TSNE_2), data = finalPlots[finalPlots$TypeVacc != "cDC1",], color = "lightgrey", size=0.5) + 
  geom_pointdensity(adjust=4, size = 1) +
  scale_color_viridis_c(limits = c(0,325), option = "magma")+
  theme_classic() + ggtitle("TSNE by cDC1")#+ 

```

# FIG. 6J - GSEA of CD4 Trm cluster compared with CD4 Tem and CD4 IL7R
```{r, fig.width=6, fig.height=3.5}
cells_C0 <- seuratSC.Scal.comb@meta.data %>% filter(
    Cell_type_Specific_1 == "CD4+ Trm") %>% pull(Cell_ID)

cells_CD4_rest <- seuratSC.Scal.comb@meta.data %>% filter(
    Cell_type_Specific_1 %in% c("CD4+ Teff", "CD4+ Teff Th17")) %>% pull(Cell_ID)

#Compare using Findmarkers
list_markers_CD4 <- FindMarkers(
      object = seuratSC.Scal.comb, ident.1 = cells_C0, ident.2 = cells_CD4_rest, 
      test.use = "wilcox", min.pct = 0.2
    )  

list_markers_CD4$gene <- rownames(list_markers_CD4)

#Prepare gene datasets 
mmuGeneSets <- getGenesets(
  org = "mmu", db = "kegg", cache = TRUE, 
  return.type = "list", gene.id.type = "SYMBOL"
)
mmuGeneSets <- lapply(X = mmuGeneSets, FUN = toupper)

#Run GSEA from ranked genes
ranks <- list_markers_CD4 %>% arrange(desc(avg_log2FC)) %>% 
    select(gene, avg_log2FC) %>% deframe()

CD4_K <- fgsea(pathways = mmuGeneSets, stats = ranks, minSize =  10, maxSize = 500)

#Plot most significant pathways 
CD4_K_sig <- rbind(top_n(CD4_K, 7, NES), top_n(CD4_K, -4, NES))
CD4_K_sig <- CD4_K_sig[order(CD4_K_sig$NES, decreasing = FALSE),]

#Delete irrelevant rows related with diseases
CD4_K_sig <-  CD4_K_sig %>% filter(pathway != "mmu05012_Parkinson_disease")
CD4_K_sig <-  CD4_K_sig %>% filter(pathway != "mmu05016_Huntington_disease")
CD4_K_sig <-  CD4_K_sig %>% filter(pathway != "mmu05020_Prion_disease")
CD4_K_sig$pathway <- factor(CD4_K_sig$pathway, levels = CD4_K_sig$pathway)

ggplot(CD4_K_sig, aes(x = NES, y = pathway, fill = pathway)) +
  geom_bar(stat = "identity") +  # Use geom_bar for barplot
  scale_fill_manual(values = c(rep("#5c8f9c", each = 4), rep("#654321", each = 4))) +
  labs(
    x = "NES",
    y = "Pathway",
    fill = "Adjusted p-value"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 2),  # Customize axis text
    axis.title.x = element_text(size = 14)
    #axis.title.y = element_text(size = 14)
  )
```

# FIG. 6K - Violin plot on selected DEG genes enriched on CD4 Trm compared with CD4 Tem and CD4 IL7R
```{r, fig.width=6, fig.height=3.5}
vgenes <- c('CCR5', 'CCR2', 'IFNG', 'TNF','TGFB1', 'CD40LG', 'PDCD1', 'LAG3', 'HAVCR2', 'TNFRSF4', 'TNFRSF9')
cluster_order = c('C0','C3','C11','C4','C7','C2','C5','C1','C8','C12', 'C6', 'C9', 'C10')

### make table for ggplot
setdiff(vgenes, rownames(seuratSC.Scal.comb))
genexp <- data.frame(t(seuratSC.Scal.comb@assays$RNA@data[vgenes, ]))
genexp <- cbind(seuratSC.Scal.comb@meta.data[,'res.Final.1'], rownames(genexp), genexp); dim(genexp)
names(genexp)[names(genexp) == 'seuratSC.Scal.comb@meta.data[, \"res.Final.1\"]'] <- 'seurat_cluster'
names(genexp)[names(genexp) == 'rownames(genexp)'] <- 'Row.names'

genexp.m <- melt(genexp, id.vars=c('Row.names','seurat_cluster'))
genexp.m$seurat_cluster <- factor(genexp.m$seurat_cluster, levels = cluster_order)
gcolors <- color.list()[1:13]
gcolors <- colorsPlots$Cell_type_Specific_1
gcolors <- unname(gcolors)


### Select only relevant clusters
genexp.m <- genexp.m %>% filter(seurat_cluster %in% c("C0", "C3", "C11"))


### plot

ggplot(genexp.m) + geom_violin(aes(x=variable, y=value, fill=seurat_cluster), color='black', scale='width', kernel='gaussian', adjust=1.5) + 
  scale_y_continuous(name="Expression", limits=c(-2, 3), breaks=c(0,3)) +
  theme(axis.ticks = element_blank()) +
  facet_grid(rows='seurat_cluster') + 
  theme_classic() + scale_fill_manual(values=gcolors) +
  theme(axis.text.x = element_text(angle = 90)) + xlab('Gene') + NoLegend() + 
  theme(panel.spacing.y = unit(0, "lines"), strip.text.y = element_blank())

```